Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
incorporate introduction comments GGI RZ

Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation. The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].

Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlithg a stronger reponse of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of species distribution and where species performance is considered to be threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use at their present climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show lower-values of resilience after several extreme drought only at the dry edge of climatic gradients. To address this hyphotesis our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change along climatic gradients. To achieve goal (*iii*) within the region of study we assessed whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.  [@Bhuyan2017].

>>>>

Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of the Mediterranean climate [@Dai2011]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation. Thus, assessment of drought impacts on ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].

Ecosystem responses to drought are influenced by other drivers of environmental changes. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. This becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.

Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018]. The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impact in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge. To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of this species to detect vulnerability to climate change along climatic gradients. To achieve goal (*iii*) we assessed within the region of study whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.  [@Bhuyan2017].
